---
title: Public Health Indicator Analysis
date: 6/25/20
output: html_notebook
---

In this project the indicator *Breast tumor removal surgery
within 60 days from the date of diagnosis* is defined. The
indicator for each hospital structure present in the data
is analyzed and the association at individual level between
the level of education and the work and the value of the indicator is
also assessed.

### Dataset description
Three simulated datasets are available, two extracts from population
registers and one from an administrative health flow:

- extract from a health registry for the year 1984
relating to a small town. The variables that compose it are t
he following:
    - **idnum**: identification code of the subject
    - **smoke** (yes or no): if the subject currently smokes
    - **sex** (Female or Male)
    - **married** (yes or no): if it is currently married
    - **kids** (yes or no): if the subject has children
    - **work** (yes or no): if it currently works
    - **education** (no / low <= middle school diploma; medium / high> = middle school diploma
    higher
    - **age** (numerical in years)

- Cancer Registry for the month of January 1984. The variables
 that make it up are the following:
    - **idnum**: identification code of the subject
    - **stage** (I, II, III, IV): stage of the tumor at diagnosis
    - **incidence**: date of diagnosis of the tumor
    - **type of tumor**: breast, lung, colon, other

- Hospital discharge sheets of subjects hospitalized between
January 1984 and October 1984 for oncological treatments:
    - **idnum**: identification code of the subject
    - **performance**: type of treatment received during hospitalization (surgical, chemotherapy or radiotherapy)
    - **date of performance**: date of treatment
    - **discharge**: date of discharge from the hospital
    - **hospital**: unique code of the hospital

### Analysis
In the first phase, the data sets are cleaned up removing
the inconsistencies, then the record linkage of the data
sets is performed. The indicator is calculated globally and
by individual healthcare facility.\
The odds ratio is calculated for both the education variable
and the job variable, using the Mantel Haenszel method.
Estimate and interpret this effect measure, adjusted for all
the available variables that you consider appropriate to
include as potential confounders, using a logistic regression
model.


### Import Libraries
```{r}
library(ggplot2)
library(Epi)
library(DescTools)
```

### Import dataset
```{r}
# import dataset
df_anagrafica <- read.csv2("DatiAnagrafici.csv", sep = ";", header = TRUE, na="")
df_tumors <- read.csv2("RegistroTumori.csv", sep = ";", header = TRUE, na="")
df_sdo <- read.csv2("SDO.csv", sep = ";", header = TRUE, na="")
```

### Point 1
Review the datasets and report descriptive statistics in a table for each dataset. Report minimum and maximum observation dates for datasets 2 and 3. Pay attention to the possibility of missing data, inconsistencies between dates, repeated records that could create problems in the linkage and analysis phase.
Records with missing or repeated data or inconsistencies between dates (eg. Treatment date prior to the incidence date) must be reported in the report and then deleted for subsequent analyzes.

Descriptive statistics are shown for each dataset.
```{r}
# Dataset anagrafica
summary(df_anagrafica)
```

```{r}
# Dataset tumori
summary(df_tumors)
```

```{r}
# Dataset SDO
summary(df_sdo)
```

A cleaning phase of the various datasets is performed, eliminating null values and inconsistencies.

#### Personal data
In the dataset there are records with missing values for the columns *married*, *kids* and *education*
that must be eliminated.
```{r}
# Record indices with missing values
missing_anagrafica <- sort(unique(c(which(is.na(df_anagrafica$married)),
                                    which(is.na(df_anagrafica$kids)),
                                    which(is.na(df_anagrafica$education)))))

# Record with missing values
df_anagrafica[missing_anagrafica,]
```

Duplicate records are checked.
```{r}
df_anagrafica[duplicated(df_anagrafica),]
```

There are no duplicate records. The clean dataframe is defined for analysis.
```{r}
df_anagrafica_clear <- df_anagrafica[-missing_anagrafica,]
```

With a total of `r nrow(df_anagrafica_clear)` record.


#### Tumors

The *incidenza* column is changed to date
```{r}
df_tumors$incidenza = as.Date(df_tumors$incidenza, "%d/%m/%Y")
```

There are missing values for the *stadio*, *tipotumore* and *incidenza* columns
```{r}
missing_tumors <- sort(unique(c(which(is.na(df_tumors$Stadio)),
                                which(is.na(df_tumors$tipotumore)),
                                which(is.na(df_tumors$incidenza)))))

# Record with missing values
df_tumors[missing_tumors,]

# Creating the dataframe without missing values
df_tumors_clear <- df_tumors[-missing_tumors,]
```

In total the number of deleted records are `r length(missing_tumors)`. Duplicate records are checked.
```{r}
# Indices of duplicate records
dup_ind <- duplicated(df_tumors_clear)

# Duplicate records are
df_tumors_clear[dup_ind,]

# Elimination of duplicates
df_tumors_clear <- unique(df_tumors_clear)
```
With a total of `r nrow(df_tumors_clear)` record.


#### SDO
The *dataprestazione* and *dimissione* columns are converted to dates.
```{r}
df_sdo$dataprestazione = as.Date(df_sdo$dataprestazione, "%d/%m/%Y")
df_sdo$dimissione = as.Date(df_sdo$dimissione, "%d/%m/%Y")
```

Missing values for the columns *dataprestazione*, *dimissione*.
```{r}
missing_sdo <- sort(unique(c(which(is.na(df_sdo$dataprestazione)),
                             which(is.na(df_sdo$dimissione)))))

# Record with missing values
df_sdo[missing_sdo,]

# Creating the dataframe without missing values
df_sdo_clear <- df_sdo[-missing_sdo,]
```

There are no duplicate records.
```{r}
# duplicate records
df_sdo_clear[duplicated(df_sdo_clear),]
```

There are inconsistencies between the dates of the service and the dates of discharge.
```{r}
date_no_ok <- df_sdo_clear$dataprestazione > df_sdo_clear$dimissione

# Records with inconsistent dates are
df_sdo_clear[date_no_ok,]

# These records are deleted
df_sdo_clear <- df_sdo_clear[-date_no_ok,]
```

With a total of `r nrow(df_sdo_clear)` record.

The minimum and maximum observation dates are calculated for the tumor and SDO datasets.
```{r}
# data for the tumor dataset
max_data_tumors <- df_tumors_clear[which.max(df_tumors_clear$incidenza), 'incidenza']
mim_data_tumors <- df_tumors_clear[which.min(df_tumors_clear$incidenza), 'incidenza']
diff_tumors <- max_data_tumors - mim_data_tumors

# performance date for the tumor dataset
max_data_sdo_dataprestazione <- df_sdo_clear[which.max(df_sdo_clear$dataprestazione),'dataprestazione']
mim_data_sdo_dataprestazione <- df_sdo_clear[which.min(df_sdo_clear$dataprestazione),'dataprestazione']
diff_sdo_dataprestazione <- max_data_sdo_dataprestazione - mim_data_sdo_dataprestazione

# discharge for the tumors dataset
max_data_sdo_dimissione <- df_sdo_clear[which.max(df_sdo_clear$dimissione),'dimissione']
mim_data_sdo_dimissione <- df_sdo_clear[which.min(df_sdo_clear$dimissione),'dimissione']
diff_sdo_dimissione <- max_data_sdo_dimissione - mim_data_sdo_dimissione

# Creating a complete dataframe
df_date <- data.frame(c("tumor incidence date", "sdo performance date", "sdo discharge date"),
                      c(mim_data_tumors, mim_data_sdo_dataprestazione, mim_data_sdo_dimissione),
                      c(max_data_tumors, max_data_sdo_dataprestazione, max_data_sdo_dimissione),
                      c(diff_tumors, diff_sdo_dataprestazione, diff_sdo_dimissione))
colnames(df_date) <- c("Dates", "Minimum date", "Maximum date", "Difference in days")

df_date
```

### Point 2
Record the linkage of the 3 datasets with the aim of building the indicator *Breast cancer removal surgery
within 60 days from the diagnosis date* on a monthly basis for the incidents in January 1984.

The linkage record of the three clean datasets is performed.
```{r}
df_merge <- merge(x = df_anagrafica_clear, y = df_tumors_clear, by = "idnum", all.x = TRUE)
df_final <- merge(x = df_merge, y = df_sdo_clear, by = "idnum", all.x = TRUE)
head(df_final)
```

It is verified that the treatment date is later than the tumor incidence date, records for which this is not true
will be removed from the dataset.
```{r}
df_final[
  !is.na(df_final$dataprestazione) &
    !is.na(df_final$incidenza) &
    df_final$dataprestazione < df_final$incidenza, ]

```

There is no record with this inconsistency.


The indicator is built on a monthly basis for incidents in January 1984.
The denominator is defined as the number of female subjects with breast cancer occurring between 01/01/1984
and 31/01/1984, in stage I or II, who have undergone surgery
```{r}
# Calculation of the denominator
den_ind <- which(
  df_final$sex == "Female" & df_final$tipotumore == "seno" &
    (df_final$Stadio == "Stadio I" | df_final$Stadio == "Stadio II") &
    df_final$incidenza <= as.Date("31/01/1984", format = "%d/%m/%Y") &
    df_final$incidenza >= as.Date("01/01/1984", format = "%d/%m/%Y") &
    df_final$Prestazione == "chirurgica")

den <- df_final[den_ind,]
nrow(den)
```

The numerator is defined as the number of all subjects in the denominator with an interval between the incidence
date and the intervention date ≤ 60 days.
```{r}
# Calculation of the denominator
num_ind <- which(den$dataprestazione - den$incidenza <= 60)

num <- den[num_ind,]

nrow(num)
```

The indicator and confidence interval are calculated with a 95% confidence level.
```{r}
ind <- as.data.frame(BinomCI(nrow(num), nrow(den), conf.level = 0.95))

ind
```

### Point 3
Calculate the indicator for hospital and give a graphic representation, including the indicator calculated on the
entire dataset as the reference value in the graph.
```{r}
num
```
The differentiated indicator per hospital is calculated
```{r}
# Grouping by hospital
df_group_osp <- merge(aggregate(num$ospedale, by=list(Category=num$ospedale), FUN=length),
                      aggregate(den$ospedale, by=list(Category=den$ospedale), FUN=length),
                      by = "Category",)

colnames(df_group_osp) <- c("Hospital", "Numerator", "Denominator")
df_group_osp$Hospital <- as.factor(df_group_osp$Hospital)

df_group_osp
```

For each hospital, the indicator and the relative confidence interval are calculated
```{r}
df_indicatore <- as.data.frame(BinomCI(df_group_osp$Numerator,
                                       df_group_osp$Denominator,
                                       conf.level = 0.95))

df_group_osp$Indicator <- df_indicatore$est
df_group_osp$lowerCI <- df_indicatore$lwr.ci
df_group_osp$upperCI <- df_indicatore$upr.ci

df_group_osp
```

The various calculated indicators are shown.
```{r}
p <- ggplot(df_group_osp, aes(x=Hospital, y=Indicator)) +
  geom_col(fill="light blue")

p + geom_text(aes(label = round(Indicator, 3)), size=6) +
  annotate("text", x = 4.5, y = 0.85, size=6, fontface="bold",
           label = paste0("General indicator: ", round(ind[1], 3),
                          " - C.I. = (", round(ind[2], 3), ", ", round(ind[3], 3), ")")) +
  geom_errorbar( aes(x=Hospital, ymin=lowerCI, ymax=upperCI), width=0.4, colour="black", alpha=0.4, size=1) +
  ylab("Indicator value") +
  theme(axis.title.x = element_text(face="bold", size=20),
        axis.text.x  = element_text(vjust=0.5, face="bold", size=16),
        axis.title.y = element_text(face="bold", size=20),
        axis.text.y  = element_text(vjust=0.5, face="bold", size=16))
```

### Point 4
Use the dataset obtained to evaluate the association on an individual level between the level of education
and the value of the indicator.

#### Quale misura di effetto è possibile stimare?

The **odds ratio** is used to calculate the association, in this way we want to understand if different levels of
education have a statistically significant impact on the calculated indicator.
```{r}
# The intervention variable is created based on the calculated indicator
den$surgery <- ifelse(rownames(den) %in% rownames(num), 1, 0)
```

The odds ratio is calculated
```{r}
t <- twoby2(den$education,den$surgery)
```
The odds ratio is `r round(t$measures[2], digit=3)`, furthermore, the confidence interval contains the value 1
and therefore the odds ratio is not statistically different from 1. The level of education therefore does not
affect the event.


#### Calculate the same effect measure, this time adjusted for the only *working* variable, using the Mantel Haenszel method.

The three-dimensional contingency table is created considering also the work variable.
```{r}
cont_table <- table(education = den$education,
                    intervento = den$surgery,
                    work = den$work)

cont_table
```

**Breslow-Day** homogeneity test is carried out
```{r}
BreslowDayTest(cont_table)
```
The **p-value** is greater than 0.05, therefore the null hypothesis of homogeneity between the groups cannot
be rejected.

**Mantel Haenszel**'s test is performed to obtain the value of his statistic, the combined p-value and odds ratio.
```{r}
mh_test <- mantelhaen.test(cont_table)
mh_test
```
The statistic p-value is greater than 0.05. The null hypothesis is not rejected, namely that the OR of the two tables
is not statistically different and the working variable does not affect the outcome.


#### Estimate and interpret this effect measure, adjusted for all the available variables that you consider appropriate to include as potential confounders, using a logistic regression model.

As possible confounding variables we consider: the age, the stage of the tumor and the hospital where the service was performed.
```{r}
model1 <- glm(surgery ~  age + factor(Stadio) + factor(ospedale),
              data = den, family = "binomial")
summary(model1)
```

Calculation of odds ratios
```{r}
exp(cbind("OR" = coef(model1), confint.default(model1, level = 0.95)))
```
As for the stage, the fact of having a second stage tumor makes the intervention more likely within 60 days of
the incidence, this is also confirmed by the confidence interval which does not contain the value 1.
For all the other variables the difference in the values assumed is not statistically different.

Another model was tried in which the age variable was categorized according to its value, three bands are
identified: the first for age up to 50 years, the second for age up to 70 and the third for age over 70
```{r}
den$age_cat <- ifelse(den$age < 50, 1, ifelse(den$age < 70, 2, 3))

model2 <- glm(surgery ~ factor(age_cat) + factor(Stadio) + factor(ospedale),
              data = den, family = "binomial")

summary(model2)
```

The related odds ratios
```{r}
exp(cbind("OR" = coef(model2), confint.default(model2, level = 0.95)))
```

No particular differences are obtained between the two models.